\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-20 doc/mlsys_paper/paper.tex
Final changes.  Also updates the definition of partially differentiable.

\begin{abstract}
\vspace*{-0.3em}
%% the abstract below still needs more meat and sharpening
We present {\tt ensmallen}, a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
Several types of optimizations are supported, including differentiable,
separable, constrained, and categorical objective functions.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and a new objective function requires typicalle one or two C++ functions. 
This can aid in the quick implementation and prototyping of new machine learning algorithms.
Due to the use of C++ template metaprogramming, {\tt ensmallen} is able to
support compiler optimizations that provide fast runtimes.
Empirical comparisons show that {\tt ensmallen} is able to outperform other
optimization frameworks (like Julia and SciPy), sometimes by large margins.
The library is distributed under the
% save words
% 3-clause
BSD license and is ready for use
in production environments.
\end{abstract}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms.
We present {\tt ensmallen}, a C++ optimization toolkit
that contains a wide variety of optimization techniques for many types of
objective functions.  Through the use of C++ template
metaprogramming~\cite{Abrahams_2004,Vandevoorde_2018},
{\tt ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

Although there are many existing machine learning optimization toolkits, few
are able to take explicit advantage of metaprogramming based code optimizations,
and few offer robust support for various types of objective functions.
For instance, deep learning
libraries like Caffe~\cite{jia2014caffe},
PyTorch~\cite{paszke2017automatic},
and TensorFlow~\cite{abadi2016tensorflow}
each contain a variety of optimization techniques.  However, these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Other machine learning
libraries, such as {\tt scikit-learn}~\cite{pedregosa2011scikit}
contain optimization algorithms but not in a coherent or reusable framework.
Many programming languages have higher-level packages for
mathematical optimization.  For example, {\tt
scipy.optimize}~\cite{jones2014scipy},
is widely used in the Python community, and MATLAB's function optimization
support has been available and used for many decades.
However, these
implementations are often unsuitable for modern machine learning tasks---for
instance, computing the full gradient of the objective function may not be
feasible because there are too many data points.

In this paper, we describe the functionality of {\tt ensmallen} and the types of
problems that it can be applied to.  We discuss the mechanisms by which {\tt ensmallen} is
able to provide both computational efficiency and ease-of-use.
We show a few examples that use the library, as well as empirical performance comparisons
with other optimization libraries.

{\tt ensmallen} is open-source software licensed under the 3-clause BSD
license~\cite{Rosen_2004_full}, 
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{sanderson2016armadillo} is used for efficient linear algebra operations,
with optional interface to GPUs via NVBLAS~\cite{nvidia2015}.

@@

        & {\tt ensmallen}            & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE \\

@@

        & Julia ({\tt Optim.jl}) \cite{mogensen2018optim}         & \CIRCLE & - & - & \CIRCLE & - & - & - \\

@@

{\tt ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the {\tt ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\argminright_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for {\tt
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  %\item {\bf numeric}: all elements of $x$ take values in $\mathcal{R}$
  \item {\bf sparse}: the gradient $f'(x)$ or $f'_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the gradient $f'_j(x)$ is computable for
individual elements $x_j$ of $x$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}
\vspace*{-0.3em}

Due to its straightforward abstraction framework, {\tt ensmallen} is able to
provide a large set of diverse optimization algorithms for objective functions
with these properties.  Below is a list of currently available optimizers:

@@

In {\tt ensmallen}'s framework, if a user wants to optimize a differentiable objective
function, they only need to provide implementations of $f(x)$ and $f'(x)$, and
then they can use any of the gradient-based optimizers that {\tt ensmallen}
provides.  Table~\ref{tab:functionality} contrasts 
the classes of objective functions that can be handled by {\tt ensmallen}
and other popular frameworks and libraries.

Not every optimization algorithm provided by {\tt ensmallen} can be
used by every class of objective function; for instance, a gradient-based
optimizer such as L-BFGS cannot operate on a non-differentiable objective
function.
As such, the best the library can attain is to maximize the flexibility
available, so that a user can easily implement a function $f(x)$ and have it
work with as many optimizers as possible.

To accomplish the flexibility, {\tt ensmallen} makes heavy use of
C++ template metaprogramming.
When implementing an objective function to be optimized,
a user may only need to implement a few methods; metaprogramming is then automatically
used to check that the given functions match the requirements of the
optimizer that is being used.  When implementing an optimizer, we can assume
that the given function to be optimized meets the required assumptions of the
optimizers, and encode those requirements as compile-time checks
(via \texttt{\small static\_assert}).

@@

We can implement these two functions in a class named {\tt LinearRegressionFunction},
as shown in Fig.~\ref{fig:LinearRegressionFunction}.
This is the entire required implementation to optimize the linear regression model with
any of the gradient-based optimizers in {\tt ensmallen}.

@@

Given the user-defined {\tt LinearRegressionFunction} class,
the code snippet below
shows how the L-BFGS optimizer can be used to find the best parameters $\theta$:

@@

To use the small-batch SGD-like optimizers provided by {\tt ensmallen},
only a slight variation on the signature of \texttt{\small Evaluate()} and
\texttt{\small Gradient()} would be needed, plus the \texttt{\small Shuffle()}
and \texttt{\small NumFunctions()} utility methods.

@@

Using template metaprogramming, {\tt ensmallen} provides an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual \texttt{\small Evaluate()} and \texttt{\small Gradient()} functions, a user
may simply write an \texttt{\small EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  As an example,
for the 
\texttt{LinearRegressionFunction} class in Fig.~\ref{fig:LinearRegressionFunction},
we can replace \texttt{\small Evaluate()} and \texttt{\small Gradient()}
with an implementation of \texttt{\small EvaluateWithGradient()}
that computes {\small $(y - X \theta)$} only once:

@@

Template metaprogramming techniques are automatically used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if \texttt{\small EvaluateWithGradient()} is not provided, a version will be
automatically generated that calls both \texttt{\small Gradient()} and \texttt{\small Evaluate()} in turn.
Similarly, if \texttt{\small Evaluate()} or \texttt{\small Gradient()} does not exist, then {\tt
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

The use of template metaprogramming in this manner also allows for compiler optimizations that
would not otherwise be possible (and that are often not possible in other
frameworks).  Firstly, because the objective function class itself is a template
parameter, the compiler is able to avoid the overhead of a function pointer
dereference, which would not be easily possible when using a language with
virtual inheritance.  The compiler is also able to use inlining and any
optimizations that may imply, including removing temporary values and dead code
elimination.  Further, if {\tt ensmallen} automatically generates an
\texttt{\small Evaluate()} or \texttt{\small Gradient()} method from a user-supplied
\texttt{\small EvaluateWithGradient()} method, the compiler can in some cases recognize and
remove the computation of unnecessary results.  For instance, in an
automatically generated \texttt{\small Evaluate()} method, the computation of the gradient
from \texttt{\small EvaluateWithGradient()} can be avoided entirely.

Overall, the automatic code generation functionality in {\tt ensmallen}
reduces the requirements for users
when they are implementing their own objective functions to be optimized,
and allows users a way to provide more efficient implementations of their
objective functions.
This leads to quicker development, quicker results, and reduces the likelihood of bugs.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & {\tt ensmallen} & {\tt scipy} & {\tt Optim.jl} & {\tt samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
default & {\bf 0.004s} & 1.069s & 0.021s & 3.173s \\
tuned & & 0.574s & & 3.122s \\ % TODO
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with various
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.  The {\it tuned} row indicates that the code was manually
modified for speed.}
\label{tab:rosenbrock_results}
\end{table}

To demonstrate the benefits of the metaprogramming based code optimizations
that {\tt ensmallen} can exploit,
we compare the performance of {\tt ensmallen} with several other
optimization frameworks, including some that use automatic differentiation.

@@

We compare four frameworks for this task:
{\bf (i)}~{\tt ensmallen},
{\bf (ii)}~\texttt{\small scipy.optimize.anneal} from SciPy 0.14.1~\cite{jones2014scipy},
{\bf (iii)}~simulated annealing implementation in \texttt{\small Optim.jl} with Julia 1.0.1~\cite{mogensen2018optim},
{\bf (iv)}~\texttt{\small samin} in the \texttt{\small optim} package for Octave~\cite{octave}.
While another option here might be \texttt{\small simulannealbnd()} 
in the Global Optimization Toolkit for MATLAB,
no license was available.
% TODO: get Marcus' system specs.
We ran our code on a MacBook Pro i7 2018 with 16GB RAM running macOS 10.14 with clang 1000.10.44.2, Julia version 1.0.1, Python 2.7.15, and Octave 4.4.1.

% We compare four frameworks%
% %
% \footnote{Another option here might be {\tt simulannealbnd()} 
% in the Global Optimization Toolkit for MATLAB.
% However, no license was available for these simulations.}
% %
% for this task:
% 
% \vspace*{-0.3em}
% \begin{itemize}
% \renewcommand{\itemsep}{-0.5ex}
%   \item {\tt ensmallen}
%   \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
%   \item simulated annealing implementation in {\tt Optim.jl} with Julia
% 1.0.1~\cite{mogensen2018optim}
%   \item {\tt samin} in the {\tt optim} package for GNU Octave~\cite{octave}
% \end{itemize}
% \vspace*{-0.3em}

Initially, we implemented these functions as simply as possible and ran them
without any tuning. This reflects how a typical user might interact with a
given framework.
The results are shown in the first row of
Table~\ref{tab:rosenbrock_results}.  Only Julia and {\tt
ensmallen} are compiled, and thus are able to avoid the function pointer
dereference and take advantage of inlining and related optimizations.

@@

% For {\tt ensmallen} we have 2 versions:
% (i)~with only \texttt{\small EvaluateWithGradient()},
% and
% (ii)~with \texttt{\small Evaluate()} and \texttt{\small Gradient()}.
% The code for these functions is as shown earlier.
% For Julia we have the options of using manually defined objective and gradient functions,
% or the gradient function can be automatically computed by 
% {\tt Calculus.jl}
% (\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl})
% or {\tt ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016}.

@@

\begin{table}
\centering
\begin{adjustbox}{scale={0.90}{0.90}}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
{\tt ensmallen}-1 & {\bf 0.001s} & {\bf 0.009s} & {\bf 0.154s} & {\bf 2.215s} \\
{\tt ensmallen}-2 & 0.002s & 0.016s & 0.182s & 2.522s \\
% Dropped for space and awful performance
%{\tt Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
{\tt Optim.jl} & 0.006s & 0.030s & 0.337s & 4.271s \\
{\tt scipy} & 0.003s & 0.017s & 0.202s & 2.729s \\
{\tt bfgsmin} & 0.071s & 0.859s & 23.220s & 2859.81s\\
% It's possible to tune ForwardDiff.jl a bit, but it doesn't give significant
% speedups to make it competitive and it really makes the code ugly.
{\tt ForwardDiff.jl} & 0.497s & 1.159s & 4.996s & 603.106s \\
{\tt autograd} & 0.007s & 0.026s & 0.210s & 2.673s \\
\bottomrule
\end{tabular}
\end{adjustbox}
\vspace*{0.25ex}
\caption{\footnotesize
Runtimes for the linear regression function on various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
%\vspace*{-1ex}
\end{table}

@@

Lastly, we demonstrate the easy pluggability in {\tt ensmallen}
for using various optimizers on the same task.
Using a version of \texttt{\small LinearRegressionFunction} from Sec.~\ref{sec:linreg_example}
adapted for separable differentiable optimizers,
we run six optimizers with default parameters in just 8 lines of code,
as shown in Fig.~\ref{fig:learning_curve}(a).
Applying these optimizers to the \texttt{\small YearPredictionMSD}
dataset from the UCI repository~\cite{ucimlrepository}
yields the learning curves shown in Fig.~\ref{fig:learning_curve}(b).
Any other optimizer for separable differentiable objective
functions can be dropped into place in the same manner.
%% Just because we have some extra space...
%% CS: we need space for the acknowledgement section
% This facilitates the seamless evaluation of various optimizers
% for user-defined objective functions.

@@

We have described {\tt ensmallen}, a flexible C++ library for function
optimization that provides an easy interface for the implementation and optimization
of user-defined objective functions.  Many types of functions can be optimized,
including separable and constrained functions.
The library provides many pre-built optimizers (including numerous variants
of SGD and Quasi-Newton optimizers).
The library internally exploits template metaprogramming
to maximise opportunities for efficiency gains,
as well as 
to make the implementation of objective functions easier
by automatically generating missing methods.

We aim to expand the library with further optimization techniques
as the need arises.  Since {\tt ensmallen} is open source,
external contributions to the codebase are welcome.
For more information on the
library, see the website and documentation at
\href{https://www.ensmallen.org}{\tt\small https://www.ensmallen.org}.
The library is already in use for function optimization in the
\texttt{\small mlpack} machine learning toolkit~\cite{mlpack2018}.

@@

%\begin{small}
{\bf Acknowledgements.}
We would like to thank the many contributors to {\tt ensmallen},
who are listed on the associated website.
%\end{small}

% \subsubsection*{Acknowledgements}
% \vspace*{-0.5em}
% 
% The development team of {\tt ensmallen} does not include just the authors named
% here but also a long list of other contributors.  See
% \url{https://www.ensmallen.org/about.html} for more information.
% % TODO: that URL may change

>>>>

\begin{abstract}
\vspace*{-0.3em}
%% the abstract below still needs more meat and sharpening
We present \texttt{\small ensmallen}, a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
Several types of optimizations are supported, including differentiable,
separable, constrained, and categorical objective functions.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and a new objective function requires typically one or two C++ functions. 
This can aid in the quick implementation and prototyping of new machine learning algorithms.
Due to the use of C++ template metaprogramming, \texttt{\small ensmallen} is able to
support compiler optimizations that provide fast runtimes.
Empirical comparisons show that \texttt{\small ensmallen} is able to outperform other
optimization frameworks (like Julia and SciPy), sometimes by large margins.
The library is distributed under the
% save words
% 3-clause
BSD license and is ready for use
in production environments.
\end{abstract}

@@

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms.
We present \texttt{\small ensmallen}, a C++ optimization toolkit
that contains a wide variety of optimization techniques for many types of
objective functions.  Through the use of C++ template
metaprogramming~\cite{Abrahams_2004,Vandevoorde_2018},
\texttt{\small ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

Although there are many existing machine learning optimization toolkits, few
are able to take explicit advantage of metaprogramming based code optimizations,
and few offer robust support for various types of objective functions.
For instance, deep learning
libraries like Caffe~\cite{jia2014caffe},
PyTorch~\cite{paszke2017automatic},
and TensorFlow~\cite{abadi2016tensorflow}
each contain a variety of optimization techniques.  However, these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Other machine learning
libraries, such as \texttt{\small scikit-learn}~\cite{pedregosa2011scikit}
contain optimization algorithms but not in a coherent or reusable framework.
Many programming languages have higher-level packages for
mathematical optimization.  For example, \texttt{\small
scipy.optimize}~\cite{jones2014scipy},
is widely used in the Python community, and MATLAB's function optimization
support has been available and used for many decades.
However, these
implementations are often unsuitable for modern machine learning tasks---for
instance, computing the full gradient of the objective function may not be
feasible because there are too many data points.

In this paper, we describe the functionality of \texttt{\small ensmallen} and the types of
problems that it can be applied to.  We discuss the mechanisms by which \texttt{\small ensmallen} is
able to provide both computational efficiency and ease-of-use.
We show a few examples that use the library, as well as empirical performance comparisons
with other optimization libraries.

\texttt{\small ensmallen} is open-source software licensed under the 3-clause BSD
license~\cite{Rosen_2004_full}, 
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{sanderson2016armadillo} is used for efficient linear algebra operations,
with optional interface to GPUs via NVBLAS~\cite{nvidia2015}.

@@

        & \texttt{\small ensmallen}            & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE & \CIRCLE \\

@@

        & Julia (\texttt{\small Optim.jl}) \cite{mogensen2018optim}         & \CIRCLE & - & - & \CIRCLE & - & - & - \\

@@

\texttt{\small ensmallen} provides a {\bf set of optimizers} for
optimizing {\bf user-defined objective functions}.  It is also easy to implement a
new optimizer in the \texttt{\small ensmallen} framework.  Overall, our goal is to provide
an easy-to-use library that can solve the problem
%\vspace*{-0.4em}
%\begin{equation}
$\argminright_{x} f(x)$
%\end{equation}
%\vspace*{-0.4em}
%\noindent
for any function $f(x)$ that takes a vector or matrix input $x$.
In most cases, $f(x)$ will have special structure; one example might be that
$f(x)$ is differentiable.  Therefore, the abstraction we have designed for \texttt{\small
ensmallen} can optionally take advantage of this structure.  For example, in
addition to $f(x)$, a user can provide an implementation of $f'(x)$, which in
turn allows first-order gradient-based optimizers to be used.  This generally
leads to significant speedups.

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\bf arbitrary}: no assumptions can be made on $f(x)$
  \item {\bf differentiable}: $f(x)$ has a computable gradient $f'(x)$
  \item {\bf separable}: $f(x)$ is a sum of individual components: $f(x) =
\sum_{i} f_i(x)$
  \item {\bf categorical}: $x$ contains elements that can only take discrete
values
  %\item {\bf numeric}: all elements of $x$ take values in $\mathcal{R}$
  \item {\bf sparse}: the gradient $f'(x)$ or $f'_i(x)$ (for a separable
function) is sparse
  \item {\bf partially differentiable}: the separable gradient $f_i'(x)$ is also
separable for a different axis $j$
  \item {\bf bounded}: $x$ is limited in the values that it can take
\end{itemize}
\vspace*{-0.3em}

Due to its straightforward abstraction framework, \texttt{\small ensmallen} is able to
provide a large set of diverse optimization algorithms for objective functions
with these properties.  Below is a list of currently available optimizers:

@@

In \texttt{\small ensmallen}'s framework, if a user wants to optimize a differentiable objective
function, they only need to provide implementations of $f(x)$ and $f'(x)$, and
then they can use any of the gradient-based optimizers that \texttt{\small ensmallen}
provides.  Table~\ref{tab:functionality} contrasts 
the classes of objective functions that can be handled by \texttt{\small ensmallen}
and other popular frameworks and libraries.

Not every optimization algorithm provided by \texttt{\small ensmallen} can be
used by every class of objective function; for instance, a gradient-based
optimizer such as L-BFGS cannot operate on a non-differentiable objective
function.
As such, the best the library can attain is to maximize the flexibility
available, so that a user can easily implement a function $f(x)$ and have it
work with as many optimizers as possible.

To accomplish the flexibility, \texttt{\small ensmallen} makes heavy use of
C++ template metaprogramming.
When implementing an objective function to be optimized,
a user may only need to implement a few methods; metaprogramming is then automatically
used to check that the given functions match the requirements of the
optimizer that is being used.  When implementing an optimizer, we can assume
that the given function to be optimized meets the required assumptions of the
optimizers, and encode those requirements as compile-time checks
(via \texttt{\small static\_assert}).

@@

We can implement these two functions in a class named \texttt{\small LinearRegressionFunction},
as shown in Fig.~\ref{fig:LinearRegressionFunction}.
This is the entire required implementation to optimize the linear regression model with
any of the gradient-based optimizers in \texttt{\small ensmallen}.

@@

Given the user-defined \texttt{\small LinearRegressionFunction} class,
the code snippet below
shows how the L-BFGS optimizer can be used to find the best parameters $\theta$:

@@

To use the small-batch SGD-like optimizers provided by \texttt{\small ensmallen},
only a slight variation on the signature of \texttt{\small Evaluate()} and
\texttt{\small Gradient()} would be needed, plus the \texttt{\small Shuffle()}
and \texttt{\small NumFunctions()} utility methods.

@@

Using template metaprogramming, \texttt{\small ensmallen} provides an easy (and
optional) way for users to avoid this extra computational overhead.  Instead of
specifying individual \texttt{\small Evaluate()} and \texttt{\small Gradient()} functions, a user
may simply write an \texttt{\small EvaluateWithGradient()} function that returns both the
objective value and the gradient value for an input $\theta$.  As an example,
for the 
\texttt{LinearRegressionFunction} class in Fig.~\ref{fig:LinearRegressionFunction},
we can replace \texttt{\small Evaluate()} and \texttt{\small Gradient()}
with an implementation of \texttt{\small EvaluateWithGradient()}
that computes {\small $(y - X \theta)$} only once:

@@

Template metaprogramming techniques are automatically used to
detect which methods exist, and a wrapper class will use suitable mix-ins in
order to provide `missing` functionality~\cite{smaragdakis2000mixin}.  For
instance, if \texttt{\small EvaluateWithGradient()} is not provided, a version will be
automatically generated that calls both \texttt{\small Gradient()} and \texttt{\small Evaluate()} in turn.
Similarly, if \texttt{\small Evaluate()} or \texttt{\small Gradient()} does not exist, then \texttt{\small
EvaluateWithGradient()} is called, and the unnecessary part of the result will
be discarded.

The use of template metaprogramming in this manner also allows for compiler optimizations that
would not otherwise be possible (and that are often not possible in other
frameworks).  Firstly, because the objective function class itself is a template
parameter, the compiler is able to avoid the overhead of a function pointer
dereference, which would not be easily possible when using a language with
virtual inheritance.  The compiler is also able to use inlining and any
optimizations that may imply, including removing temporary values and dead code
elimination.  Further, if \texttt{\small ensmallen} automatically generates an
\texttt{\small Evaluate()} or \texttt{\small Gradient()} method from a user-supplied
\texttt{\small EvaluateWithGradient()} method, the compiler can in some cases recognize and
remove the computation of unnecessary results.  For instance, in an
automatically generated \texttt{\small Evaluate()} method, the computation of the gradient
from \texttt{\small EvaluateWithGradient()} can be avoided entirely.

Overall, the automatic code generation functionality in \texttt{\small ensmallen}
reduces the requirements for users
when they are implementing their own objective functions to be optimized,
and allows users a way to provide more efficient implementations of their
objective functions.
This leads to quicker development, quicker results, and reduces the likelihood of bugs.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & \texttt{\small ensmallen} & \texttt{\small scipy} & \texttt{\small Optim.jl} & \texttt{\small samin} \\
\midrule
% TODO: these are just single-run results from Marcus' laptop!  We need to do
% 10 and average.
default & {\bf 0.004s} & 1.069s & 0.021s & 3.173s \\
tuned & & 0.574s & & 3.122s \\ % TODO
\bottomrule
\end{tabular}
\end{center}
\caption{Runtimes for $100000$ iterations of simulated annealing with various
frameworks on the simple Rosenbrock function.  Julia code runs do not count
compilation time.  The {\it tuned} row indicates that the code was manually
modified for speed.}
\label{tab:rosenbrock_results}
\end{table}

To demonstrate the benefits of the metaprogramming based code optimizations
that \texttt{\small ensmallen} can exploit,
we compare the performance of \texttt{\small ensmallen} with several other
optimization frameworks, including some that use automatic differentiation.

@@

We compare four frameworks for this task:
{\bf (i)}~\texttt{\small ensmallen},
{\bf (ii)}~\texttt{\small scipy.optimize.anneal} from SciPy 0.14.1~\cite{jones2014scipy},
{\bf (iii)}~simulated annealing implementation in \texttt{\small Optim.jl} with Julia 1.0.1~\cite{mogensen2018optim},
{\bf (iv)}~\texttt{\small samin} in the \texttt{\small optim} package for Octave~\cite{octave}.
While another option here might be \texttt{\small simulannealbnd()} 
in the Global Optimization Toolkit for MATLAB,
no license was available.
% TODO: get Marcus' system specs.
We ran our code on a MacBook Pro i7 2018 with 16GB RAM running macOS 10.14 with clang 1000.10.44.2, Julia version 1.0.1, Python 2.7.15, and Octave 4.4.1.

% We compare four frameworks%
% %
% \footnote{Another option here might be \texttt{\small simulannealbnd()} 
% in the Global Optimization Toolkit for MATLAB.
% However, no license was available for these simulations.}
% %
% for this task:
% 
% \vspace*{-0.3em}
% \begin{itemize}
% \renewcommand{\itemsep}{-0.5ex}
%   \item \texttt{\small ensmallen}
%   \item \texttt{\small scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
%   \item simulated annealing implementation in \texttt{\small Optim.jl} with Julia
% 1.0.1~\cite{mogensen2018optim}
%   \item \texttt{\small samin} in the \texttt{\small optim} package for GNU Octave~\cite{octave}
% \end{itemize}
% \vspace*{-0.3em}

Initially, we implemented these functions as simply as possible and ran them
without any tuning. This reflects how a typical user might interact with a
given framework.
The results are shown in the first row of
Table~\ref{tab:rosenbrock_results}.  Only Julia and \texttt{\small
ensmallen} are compiled, and thus are able to avoid the function pointer
dereference and take advantage of inlining and related optimizations.

@@

% For \texttt{\small ensmallen} we have 2 versions:
% (i)~with only \texttt{\small EvaluateWithGradient()},
% and
% (ii)~with \texttt{\small Evaluate()} and \texttt{\small Gradient()}.
% The code for these functions is as shown earlier.
% For Julia we have the options of using manually defined objective and gradient functions,
% or the gradient function can be automatically computed by 
% \texttt{\small Calculus.jl}
% (\href{https://github.com/JuliaMath/Calculus.jl}{\footnotesize github.com/JuliaMath/Calculus.jl})
% or \texttt{\small ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016}.

@@

\begin{table}
\centering
\begin{adjustbox}{scale={0.90}{0.90}}
\begin{tabular}{lccccc}
\toprule
{\em algorithm} & $d$: 100, $n$: 1k & $d$: 100, $n$: 10k & $d$: 100, $n$:
100k & $d$: 1k, $n$: 100k \\
\midrule
% TODO: this was only one trial on Ryan's desktop!
\texttt{\small ensmallen}-1 & {\bf 0.001s} & {\bf 0.009s} & {\bf 0.154s} & {\bf 2.215s} \\
\texttt{\small ensmallen}-2 & 0.002s & 0.016s & 0.182s & 2.522s \\
% Dropped for space and awful performance
%\texttt{\small Calculus.jl} & 0.172s & 0.960s & 27.428s & 2535.507s \\
\texttt{\small Optim.jl} & 0.006s & 0.030s & 0.337s & 4.271s \\
\texttt{\small scipy} & 0.003s & 0.017s & 0.202s & 2.729s \\
\texttt{\small bfgsmin} & 0.071s & 0.859s & 23.220s & 2859.81s\\
% It's possible to tune ForwardDiff.jl a bit, but it doesn't give significant
% speedups to make it competitive and it really makes the code ugly.
\texttt{\small ForwardDiff.jl} & 0.497s & 1.159s & 4.996s & 603.106s \\
\texttt{\small autograd} & 0.007s & 0.026s & 0.210s & 2.673s \\
\bottomrule
\end{tabular}
\end{adjustbox}
\vspace*{0.25ex}
\caption{\footnotesize
Runtimes for the linear regression function on various dataset sizes,
with $n$ indicating the number of samples,
and $d$ indicating the dimensionality of each sample.
All Julia runs do not count compilation time.}
\label{tab:lbfgs}
%\vspace*{-1ex}
\end{table}

@@

Lastly, we demonstrate the easy pluggability in \texttt{\small ensmallen}
for using various optimizers on the same task.
Using a version of \texttt{\small LinearRegressionFunction} from Sec.~\ref{sec:linreg_example}
adapted for separable differentiable optimizers,
we run six optimizers with default parameters in just 8 lines of code,
as shown in Fig.~\ref{fig:learning_curve}(a).
Applying these optimizers to the \texttt{\small YearPredictionMSD}
dataset from the UCI repository~\cite{ucimlrepository}
yields the learning curves shown in Fig.~\ref{fig:learning_curve}(b).
Any other optimizer for separable differentiable objective
functions can be dropped into place in the same manner.
%% Just because we have some extra space...
%% CS: we need space for the acknowledgement section
% This facilitates the seamless evaluation of various optimizers
% for user-defined objective functions.

@@

We have described \texttt{\small ensmallen}, a flexible C++ library for function
optimization that provides an easy interface for the implementation and optimization
of user-defined objective functions.  Many types of functions can be optimized,
including separable and constrained functions.
The library provides many pre-built optimizers (including numerous variants
of SGD and Quasi-Newton optimizers).
The library internally exploits template metaprogramming
to maximise opportunities for efficiency gains,
as well as 
to make the implementation of objective functions easier
by automatically generating missing methods.

We aim to expand the library with further optimization techniques
as the need arises.  Since \texttt{\small ensmallen} is open source,
external contributions to the codebase are welcome.
For more information on the
library, see the website and documentation at
\href{https://www.ensmallen.org}\texttt{\small https://www.ensmallen.org}.
The library is already in use for function optimization in the
\texttt{\small mlpack} machine learning toolkit~\cite{mlpack2018}.

@@

%\begin{small}
{\bf Acknowledgements.}
We would like to thank the many contributors to \texttt{\small ensmallen},
who are listed on the associated website.
%\end{small}

% \subsubsection*{Acknowledgements}
% \vspace*{-0.5em}
% 
% The development team of \texttt{\small ensmallen} does not include just the authors named
% here but also a long list of other contributors.  See
% \url{https://www.ensmallen.org/about.html} for more information.
% % TODO: that URL may change
